Modelling last glacial cycle ice dynamics in the Alps

### 2018-07-30 articles/cycle/alps-cycle.tex
Implement comments from Susan.

    Only few regional proxy records exist that extend over periods when the Alps
    were glaciated \citep{Heiri.etal.2014}. These include lake sediment records
    in north and west of the Alps \citep[e.g.,]{Beaulieu.Reille.1992,
    Wohlfarth.etal.2008, Duprat-Oualid.etal.2017}, and cave speleothem in the
    Eastern \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
    \citep{Luetscher.etal.2015} Alps. Due to the scarcity of vegetation north
    of the Alps during glacial periods, varying sources for moisture advection,
    and the limited duration of the records, quantitative palaeoclimatic
    interpretation will require combining multiple proxies in space and time,
    and comparing them against regional circulation model output
    \citep{Heiri.etal.2014}.

@@

    Geological data from the best documented Alpine piedmont lobes indicate
    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
    potentially re-advanced to within close reach of this maximum extent, until
    as late as 22 to 17\,ka \citep[Fig.~5]{Ivy-Ochs.2015, Wirsig.etal.2016}.
    The EPICA simulations yield an
    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
    reductions) followed by a retreat and then a standstill until 17.3\,ka
    (Fig.~\ref{fig:timeseries}b, red curves) in a very good agreement with
    these geological data. On the other hand, the simulations forced by the
    GRIP palaeo-temperature record yield two distinct total ice volume maxima
    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat
    several thousand years after the dated geological evidence.

    Finally, all simulations yield very strong ice volume variability over the
    last 120\,ka (Fig.~\ref{fig:timeseries}b), indicative of many more than the
    two or three documented cycles of glacier advance and retreats onto the
    foreland for this time period \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    Palaeo-precipitation reductions dampen some of the small-scale variability,
    but they have little effect on the millennial scales that characterise
    those cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In
    particular, strong millennial variability recorded in GRIP
    \chem{\delta^{18}O} (Dansgaard-Oeschger events) have large repercussions
    on the modelled Alpine ice volume, including around six glaciations of LGM
    magnitude (Fig.~\ref{fig:timeseries}b, blue curves), which is not supported
    by geologic evidence \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    Dansgaard-Oeschger events, however, have been recorded in Central European
    lake sediments \citep{Wohlfarth.etal.2008} and cave speleothems
    \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
    their magnitude in Central Europe was likely smaller than that recorded in
    GRIP \chem{\delta^{18}O}, which either implies smaller associated
    palaeo-temperature fluctuations in Central Europe than Greenland, or
    temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
    the EPICA palaeo-temperature record has the smallest variability among the
    records
    used (Fig.~\ref{fig:timeseries}a, red curves). This results in smaller
    variability in the total ice volume, and more restrictive glaciations
    during MIS~5 and~3, (Fig.~\ref{fig:timeseries}b, red curves).

@@

    The LGM extent of Alpine glaciers has been mapped with varying level of
    detail across the Western \citep{Coutterand.2010, Buoncristiani.Campy.2011,
    Jackli.1962, Bini.etal.2009, Hantke.2011}, and Eastern Alps
    \citep{Penck.Bruckner.1909, Castiglioni.1940, Husen.1987, BGR.2007,
    Husen.2011, Bavec.Verbic.2011}. Some of these maps have been compiled
    into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
    and reproduced here (Fig.~\ref{fig:lgmvel}, red line) for comparison
    against model results.

@@

    The palaeo-climate forcing was adapted (Sect.~\ref{sec:paltemp})
    to model the maximum configuration of the Rhine Glacier in broad agreement
    with the geological
    reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
    yet discrepancies remain elsewhere. As already outlined for low resolution
    runs (Sect.~\ref{sec:footprints}), the extent of glaciation in the
    north-western Alps, including the Rhone Glacier complex, the Jura ice cap,
    and the Lyon Lobe, is underestimated in the model results
    \citep[Fig.~\ref{fig:lgmvel}a, cf][]{Bini.etal.2009, Coutterand.2010}.
    On the other hand, the model yields excessive ice
    cover in the eastern Alps, where the Mur, Drava and Sava Glaciers are
    modelled to extend tens of km beyond the mapped ice limits
    \citep[Fig.~\ref{fig:lgmvel}a, cf][]{Husen.1987, Bavec.Verbic.2011}.
    As previously discussed, these discrepancies might indicate that the
    LGM climate was characterized by an east-west gradient in temperature
    \citep[cf.][]{Heyman.etal.2013} or precipitation
    \citep[cf.][]{Wu.etal.2007} anomalies relative to present, or an east-west
    gradient in variables that are not accounted for by the PDD model.

@@

    In the Western Alps, major transfluences occur for instance across Col de
    Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across Col de Montgenèvre
    was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
    southerly ice flow across Col du Mont-Cenis has been recognized
    \citep[Fig.~3.18, p.~284]{Onde.1938, Coutterand.2010}. Similarly,
    tranfluences have been previously identified from the geomorphology across
    Simplon Pass \citep{Kelly.etal.2004} and from the distribution of glacial
    erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
    simplified climate forcing used in this sumulation could not reproduced
    the transport of glacial erratics from southern Valais to observed
    deposition sites in the Solothurn region \citep[cf.~][]{Jouvet.etal.2017a}.

@@

    Despite the low temperature variability of the palaeo-climate forcing, and
    reduced precipitation dampening glacier response, our simulation depicts
    the Alpine ice complex as highly dynamic, with many more than two or three
    \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
    glacier dynamics controlled by spatial variations in catchment size and
    hypsometry. Importantly, Dansgaard-Oeschger events, recorded in Central
    Europe \citep{Spotl.Mangini.2002, Wohlfarth.etal.2008,
    Luetscher.etal.2015} but absent from the EPICA palaeo-temperature forcing,
    may have induced an even more dynamic glacier response than that modelled
    here.

@@

    In this study the numerical ice sheet model PISM (Sect.~\ref{sec:model})
    has been applied to ice dynamics of the last glacial cycle in the Alps.
    Using three different palaeo-temperature forcing records (GRIP, EPICA, and
    MD01-2444; Sect.~\ref{sec:paltemp}), scaled to reproduce the Rhine Glacier
    piedmont lobe in agreement with the mapped LGM ice margin, and two
    different palaeo-precipitation parametrizations (with and without
    precipitation reductions; Sect.~\ref{sec:palprec}), we find that only the
    EPICA palaeo-temperature record yields model results in agreement with
    geological findings, in the sense that:

@@

    This suprising result is likely coincidental as there is no direct link
    between European and Antarctic climate. This highlights the need for more
    quantitative reconstructions of European palaeoclimate.

>>>>

    Only few regional proxy records exist that extend over periods when the Alps
    were glaciated \citep{Heiri.etal.2014}. These include lake sediment records
    in north and west of the Alps \citep[e.g.,]{Beaulieu.Reille.1992,
    Wohlfarth.etal.2008, Duprat-Oualid.etal.2017}, and cave speleothems in the
    Eastern \citep[e.g.,][]{Spotl.Mangini.2002, Boch.etal.2011} and Western
    \citep{Luetscher.etal.2015} Alps. Due to the scarcity of vegetation north
    of the Alps during glacial periods, varying sources for moisture advection,
    and the limited duration of the records, quantitative palaeoclimatic
    interpretation will require combining multiple proxies in space and time,
    and comparing them against regional circulation model output
    \citep{Heiri.etal.2014}.

@@

    Geological data from the best documented Alpine piedmont lobes indicate
    that their maximum extension during MIS~2 occurred at around 25.5--24.0\,ka
    \citep{Monegato.etal.2017}, after which Alpine glaciers remained or
    potentially re-advanced to within close reach of this maximum extent, until
    as late as 22 to 17\,ka \citep[Fig.~5]{Ivy-Ochs.2015, Wirsig.etal.2016}.
    The EPICA simulations yield an
    early maximum ice volume at 25.2/24.6 (without/with palaeo-precipitation
    reductions) followed by a retreat and then a standstill until 17.3\,ka
    (Fig.~\ref{fig:timeseries}b, red curves) in a very good agreement with
    these geological data. On the other hand, the simulations forced by the
    GRIP palaeo-temperature record yield two distinct total ice volume maxima
    at 27.3/27.0 and 21.8/21.7\,ka followed by an early deglaciation of the
    foreland by 21.4\,ka. The MD01-2444 palaeo-temperature forcing yields
    a late LGM peak ice volume at 16.5/15.7\,ka followed by a rapid retreat,
    in contrast to dated records of large-scale ice retreat by 17 to
    16\,ka\citep{Ivy-Ochs.etal.2006, Ivy-Ochs.etal.2008}.

    Finally, all simulations yield very strong ice volume variability over the
    last 120\,ka (Fig.~\ref{fig:timeseries}b), indicative of many more than the
    two or three documented cycles of glacier advance and retreats onto the
    foreland for this time period \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    Palaeo-precipitation reductions dampen some of the small-scale variability,
    but they have little effect on the millennial scales that characterise
    those cycles (Fig.~\ref{fig:timeseries}b, light colour curves). In
    particular, strong millennial variability recorded in GRIP
    \chem{\delta^{18}O} (Dansgaard-Oeschger events) have large repercussions
    on the modelled Alpine ice volume, including around six glaciations of LGM
    magnitude (Fig.~\ref{fig:timeseries}b, blue curves), which is not supported
    by geologic evidence \citep{Preusser.2004, Ivy-Ochs.etal.2008}.
    Dansgaard-Oeschger events, however, have been recorded in European
    lake sediments \citep{Wohlfarth.etal.2008} and cave speleothems
    \citep{Spotl.Mangini.2002, Luetscher.etal.2015}. Our results indicate that
    their magnitude in Europe was likely smaller than that recorded in
    GRIP \chem{\delta^{18}O}, which either implies smaller associated
    palaeo-temperature fluctuations in Europe than Greenland, or
    temperature-independent controls on GRIP \chem{\delta^{18}O}. In contrast,
    the EPICA palaeo-temperature record has the smallest variability among the
    records
    used (Fig.~\ref{fig:timeseries}a, red curves). This results in smaller
    variability in the total ice volume, and more restrictive glaciations
    during MIS~5 and~3, (Fig.~\ref{fig:timeseries}b, red curves).

@@

    The LGM extent of Alpine glaciers has been mapped with varying level of
    detail across the Western \citep{Jackli.1962, Bini.etal.2009,
    Coutterand.2010, Buoncristiani.Campy.2011, Hantke.2011}, and Eastern Alps
    \citep{Penck.Bruckner.1909, Castiglioni.1940, Husen.1987, BGR.2007,
    Husen.2011, Bavec.Verbic.2011}. Some of these maps have been compiled
    into a reconstruction, covering the entire Alps \citep{Ehlers.etal.2011},
    and reproduced here (Fig.~\ref{fig:lgmvel}, red line) for comparison
    against model results.

@@

    The palaeo-climate forcing was adapted (Sect.~\ref{sec:paltemp})
    to model the maximum configuration of the Rhine Glacier in broad agreement
    with the geological
    reconstruction of the LGM extent (Fig.~\ref{fig:lgmvel}a, solid red line),
    yet discrepancies remain elsewhere. As already outlined for low resolution
    runs (Sect.~\ref{sec:footprints}), the extent of glaciation in the
    north-western Alps, including the Rhone Glacier complex, the Jura ice cap,
    and the Lyon Lobe, is underestimated in the model results
    \citep[Fig.~\ref{fig:lgmvel}a, cf][]{Bini.etal.2009, Coutterand.2010}.
    On the other hand, the model yields excessive ice
    cover in the Eastern Alps, where the Mur, Drava and Sava Glaciers are
    modelled to extend tens of km beyond the mapped ice limits
    \citep[Fig.~\ref{fig:lgmvel}a, cf][]{Husen.1987, Bavec.Verbic.2011}.
    As previously discussed, these discrepancies might indicate that the
    LGM climate was characterized by an east-west gradient in temperature
    \citep[cf.][]{Heyman.etal.2013} or precipitation
    \citep[cf.][]{Wu.etal.2007} anomalies relative to present, or an east-west
    gradient in variables that are not accounted for by the PDD model.

@@

    In the Western Alps, major transfluences occur for instance across Col de
    Montgenèvre, Col du Mont-Cenis, Simplon Pass and Brünig Pass
    (Fig.~\ref{fig:lgmvel}a). Although a transfluence across Col de Montgenèvre
    was previously questioned \citep[Fig.~2]{Cossart.etal.2012}, evidence for
    southerly ice flow across Col du Mont-Cenis has been recognized
    \citep[Fig.~3.18, p.~284]{Onde.1938, Coutterand.2010}. Similarly,
    tranfluences have been previously identified from the geomorphology across
    Simplon Pass \citep{Kelly.etal.2004} and from the distribution of glacial
    erratics across Brünig Pass \citep{Jackli.1962}. On the other hand, the
    simplified climate forcing used in this sumulation could not reproduce
    the transport of glacial erratics from southern Valais to observed
    deposition sites in the Solothurn region \citep[cf.~][]{Jouvet.etal.2017a}.

@@

    Despite the low temperature variability of the palaeo-climate forcing, and
    reduced precipitation dampening glacier response, our simulation depicts
    the Alpine ice complex as highly dynamic, with many more than two or three
    \citep[cf.][]{Preusser.2004,Ivy-Ochs.etal.2008} glaciations, and regional
    glacier dynamics controlled by spatial variations in catchment size and
    hypsometry. Importantly, Dansgaard-Oeschger events, recorded in
    Europe \citep{Spotl.Mangini.2002, Wohlfarth.etal.2008,
    Luetscher.etal.2015} but absent from the EPICA palaeo-temperature forcing,
    may have induced an even more dynamic glacier response than that modelled
    here.

@@

    In this study the numerical ice sheet model PISM (Sect.~\ref{sec:model})
    has been applied to ice dynamics of the last glacial cycle in the Alps.
    Using three different palaeo-temperature forcing records (GRIP, EPICA, and
    MD01-2444; Sect.~\ref{sec:paltemp}), scaled to reproduce the Rhine Glacier
    piedmont lobe in agreement with the mapped LGM ice margin, and two
    different palaeo-precipitation parametrizations (with and without
    ca.~-68\,\unit{mm\,{\degree}C^{-1}}
    precipitation reductions; Sect.~\ref{sec:palprec}), we find that only the
    EPICA palaeo-temperature record yields model results in agreement with
    geological findings, in the sense that:

@@

    This interesting result may be coincidental as there is no direct link
    between European and Antarctic climate. This highlights the need for more
    quantitative reconstructions of European palaeoclimate.
